Large-scale feature selection with Gaussian mixture models for remote sensing images

### 2016-10-13 paper.tex
relecture globale

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Operational Feature Selection with Gaussian Mixture Models for remote sensing images}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

\IEEEPARstart{W}{ith} the increasing number of remote sensing missions, the quantity of available data for a given landscape becomes larger and larger. Satellite missions produce a huge amount of data on a regularly (daily) basis. From 2018, the EnMAP (Environmental Mapping and Analysis Program) satellites managed by the German space agency will produce images with 244 spectral bands with a resolution of 30x30m with a frequency of revisit of 4 days\cite{Müller09enmap}. The Hyperspectral Infrared Imager (HyspIRI) of NASA will also deliver images with 212 spectral bands. Additionally to hyperspectral data, the amount of available hypertemporal data increases a lot. For instance, the European satellites Sentinel-2 were launched successfully recently and 2 Terabytes of data will be released every days~\cite{drusch2012sentinel}. This high volume of Earth observation data provides an accurate information of land state and functioning, and help to improve the understanding of the Earth~\cite{rs1010001}. However, processing such data is more and more challenging because of statistical and computational issues.

In the spectral or temporal domain, a pixel is represented by a vector
for   which  each   component   corresponds   to  a   spectral/temporal
measurement.  The  size of  the vector  is the  number of  spectral or
temporal  measurements.   For  hyperspectral images,  this  number  is
typically   about   several   hundreds  while   for   the   Sentinel-2
multitemporal images, the number of spectro-temporal measurement for a
given  year  is approximately  one  thousand.   When working  in  high
dimensional  space,  statistical  methods  made for  low  or  moderate
dimensional  space do  not  adapt  well.  For  instance,  the rate  of
convergence of the statistical estimation decreases when the dimension
grows while conjointly the number of parameters to estimate increases,
making    the    estimation    of   the    model    parameters    very
difficult~\cite{donoho}.  Consequently,  with a limited  training set,
beyond a certain limit, the classification accuracy actually decreases
as the number of features increases~\cite{hughes}.  For the purpose of
classification,  these  problems are  related  to  the \emph{curse  of
  dimensionality}. This is a major drawback in remote sensing since it
is  difficult to  collect a  large and  an accurate  ground-truth.  An
intensive work has  been performed in the remote  sensing community to
build  accurate classifiers  for  high  dimensional images.   Bayesian
models~\cite{book:landgrebe}, feature extraction and feature reduction
techniques~\cite{book:landgrebe,DR:guided:tour},                random
forest~\cite{1396322},   neural  networks~\cite{5411821}   and  kernel
methods~\cite{kernel:methods:rs}  have   been  investigated   for  the
classification of such images.

The  size of  the data  is increasing  with respect  to the  number of
measurement per  pixel. The data  volume of an hyperspectral  image is
typically    several    Gigabytes     per    acquisition    ($\approx$
300km$^2$). Multitemporal data are  now available freely from internet
stream    (see    for    instance     the    Copernicus    data    hub
\url{https://cophub.copernicus.eu/}).  This very  large volume of data
requires   specific   computing  infrastructure.    High   performance
computing   is   actually   investigated   by   the   remote   sensing
community~\cite{christophe2011remote,plaza2011high}.  Main  issues are
related to the use of  parallel approach (multi-core, CPU, clusters) to
improve the  processing time, and  to the use of  streaming techniques
when data  does not fit in  memory.  One popular open  source software
solution  is  the Orfeo  Toolbox,  developed  by the  French  National
Agency~\cite{christophe2008orfeo}.   Streaming and  parallel computing
are  conveniently  proposed  to  the  user  through  several  modules
``\emph{ready to use}''.

A method  to reduce  both statistical and  computational issues  is to
perform a reduction of the dimension. In fact, with the \emph{curse of
  dimensionality}     comes     the      \emph{blessing     of     the
  dimensionality}~\cite{bouveyron2014model}:  high   dimensional  data
space   exhibits  interesting   properties   for  the   classification
purpose. In  particular, it is possible  to extract a set  of relevant
features    to   get    a   parsimonious    representation   of    the
data~\cite{jimenez1998supervised}.    For   instance,  in   land-cover
classification,  given  a  set   of  spatial,  temporal  and  spectral
features,  it  is  possible  to  extract  those  which  are  the  most
discriminant      for      the     purpose      of      classification
\cite{fassnacht2014comparison}.   In  hyperspectral   data,  from  the
hundreds of available spectral channels,  it is possible to reduce the
number of channels  to make the processing more efficient  in terms of
statistical complexity  and computational load. In  short, by reducing
the dimension better classification results is expected with a reduced
computational load.

@@

There are two ways to reduce dimension~\cite[Chapter 1]{Guyon:2006:FEF:1208773}:
features extraction and feature selection. Feature extraction means reformulate
and summarize the information by creating new features in combining the existing
ones, it is sometimes referred to \emph{feature construction}. Linear
combination of the initial features can be extracted using Principal Component
Analysis (PCA)~\cite{jimenez1998supervised} or Independent Component Analysis
\cite{villa2011hyperspectral}. Supervised extraction method has also been
investigated such as Fischer discriminant analysis and decision boundary feature
extraction~\cite{book:landgrebe}. To the contrary, features selection selects a
subset of existing features identified as the most relevant for a given
criterion. It has the additional advantage to be much more easily understandable for the end-user.

Features selection/extraction algorithms can be divided into three classes. The first class is called \emph{filter methods}. They select features independently of the classifier. Features are ranked according to some statistical measures (correlation, independence). For example, the PCA is a typical unsupervised filter method. Bruzzone \emph{et al.}\cite{bruzzone1995extension} develop a supervised filter method based on Jeffries-Matusita distance to maximize the separability of class distribution. The use of correlation between bands has been explored with hyperspectral data\cite{demir2008phase}. In general, these methods have the advantage to be fast and usable with any classifier but they do not take into account the strength and weakness of the chosen classifier and do not optimize directly the classification process.

The second class are known as \emph{wrapper methods}. They search for the best subset of variables for a given learning model. Since exhaustive searches are too expensive in terms of processing time, several sub-optimal search strategies have been designed, mainly iterative forward or backward search~\cite{whitney1971direct,marill1963effectiveness} or a combination of both~\cite{somol1999adaptive}. The advantage of such methods compare to filter methods is that they are dedicated to a particular model and to a particular learning problem. On the other hand, as they require the training of multiple models to test various set of variables, they tend to be more time consuming.

@@

The third class corresponds to the \emph{embedded method}. They do not separate the features selection process from the learning algorithm and allow interaction between the two processes. The basic example of such method is the decision tree algorithm in which a feature is selected for the creation of each node. Embedded methods also exist for other models, e.g. SVM \cite{guyon2002gene,weston2003use,tuia2015multiclass}.

Despite a large diversity of methods, feature selection algorithms usually do not scale well with the number of pixels to be processed~\cite{fauvel2015fast}. The training computational load is too important to compensate the reduced prediction computational load. Hence, feature selection is not so used in operational situation. However, methods based on Gaussian Mixture Models (GMM) have several interesting properties that make them suitable for feature selection in the context of large amount of data. By taking advantage of their intrinsic properties, it is possible to increase the computational efficiency with respect to standard implementation.

The contribution of the paper is a forward feature selection method using GMM in continuation of~\cite{fauvel2015fast}. An efficient implementation of the method, using linear algebra on block matrix, is presented in order to handle large amount of data. Furthermore, a floating version of the algorithm is proposed. Several criteria are proposed to handle unbalanced training set.  Finally, the developed algorithm is made available as a remote module of the C++ Orfeo Toolbox~\cite{christophe2008orfeo}.

@@

The remaining of this article is organized as follows. Section~\ref{sec:gmm-hd} presents GMM classifiers and the problems related to high-dimensional feature space. The features selection methods are detailed in Section~\ref{sec:selection}. The efficient implementation of the proposed selection method is presented in Section~\ref{sec:implementation}. Experimental results on two real high dimensional data set are given Section~\ref{sec:test}. Conclusion and perspectives conclude the paper in Section~\ref{sec:conclusion}.

@@

The    following    notations    are   used    in    the    remaining.
$\mathcal{S}  = \{\mathbf{x}_i,y_i\}_{i=1}^{n}$  denotes the  training
set where $\mathbf{x}_i \in \mathbb{R}^d$ is the vector of features of
the $i^{th}$ sample, $y_i = 1,...,C$ the associated label, C the total
number of classes,  $n$ the number of samples and  $n_c$ the number of
samples of class $c$.

@@

    For mixture models, it is assumed  that a given sample $\mathbf{x}$ is
    the realization  of a  random vector which  distribution is  a mixture
    (convex     combination)     of      several     class     conditioned
    distribution~\cite{Fraley00model-basedclustering}:
    \begin{equation}
        p(\mathbf{x}) = \sum_{c=1}^{C} \pi_c f_c(\mathbf{x}|\theta)
    \end{equation}
    where $\pi_c$ is  the prior, \emph{i.e.}, the  proportion of class
    $c$  and  $f_c$  a  parametric density  function  control by
    $\theta$.

    Among the possible parametric model,  the Gaussian one is the most
    used~\cite{bouveyron2014model}.   It assumes  that each  $f_c$ is,
    conditionally  to  $c$,  a  Gaussian  distribution  of  parameters
    $\boldsymbol{\mu}_c$    and    $\boldsymbol{\Sigma}_c$:
    \begin{equation*}
        f_c(\mathbf{x}|\boldsymbol{\mu}_c, \boldsymbol{\Sigma}_c) = \frac{1}{(2\pi)^{\frac{d}{2}} |\boldsymbol{\Sigma}_c|^{\frac{1}{2}}} \exp \left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \right).
    \end{equation*}

    It  is  referred  to  as  Gaussian  mixture  model  (GMM).   In  a
    supervised    learning    framework,    the    class    parameters
    $\boldsymbol{\mu}_c$,   $\boldsymbol{\Sigma}_c$   and  the   prior
    $\pi_c$ are  usually estimated  through the  conventional unbiased
    empirical estimators:
    \begin{align}
        \hat{\pi}_c &= \frac{n_c}{n},\\
        \hat{\boldsymbol{\mu}}_c &= \frac{1}{n_c} \sum_{\{i|y_i = c\}} \mathbf{x}_i ,\\
        \hat{\boldsymbol{\Sigma}}_c &= \frac{1}{(n_c - 1)} \sum_{\{i|y_i = c\}} (\mathbf{x}_i - \boldsymbol{\mu}_c) (\boldsymbol{}x_i - \boldsymbol{\mu}_c)^t.
    \end{align}
    To predict the  class of a new unseen sample,  the maximum \emph{a
      posteriori}  rule  is  used:
    \begin{equation*}
        \mathbf{x} \text{ belongs to } c \Leftrightarrow c = \text{arg} \max_{c \in C} p(c) p(\mathbf{x}|c).
    \end{equation*}
    Under the GMM,  and identifying $p(c)$ as $\pi_c$  and $p(\mathbf{x}|c)$ as
    $f_c(\mathbf{x}|\theta)$ and by taking the log, a decision function is obtained
    \begin{eqnarray}\label{eq:decision}
      Q_c(\mathbf{x}) &=& 2 \log \left( p(c) p(\mathbf{x}|c) \right) \nonumber \\
                      &=& - (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \nonumber \\
                      & &-\log (|\boldsymbol{\Sigma}_c|) + 2 \log (\pi_c) - d \log (2\pi).
    \end{eqnarray}

@@

    The computation of eq.~(\ref{eq:decision}) requires the inversion of the
    covariance matrix and the  computation of the logarithm of the
    determinant.  The estimation of these terms suffers from the curse
    of  dimensionality\cite{bouveyron2014model}, because  in practice,
    the  number of  parameters  $\eta_c$ to  estimate  for each  class
    increases quadratically with respect to the number of features, as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation  $n_c$ is  small  compared to  the  number of parameters
    $\eta_c$, the  estimated covariance  matrix is badly  conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.   The worst  situation  is $n_c<\eta_c$  which leads  to
    a singular covariance matrix.  Unfortunately, this situation happens
    regularly in remote sensing.   For instance in hyperspectral image
    classification,  very few  labeled samples  are usually  available
    because of the difficulty and the high cost to collect ground-truth.

@@

    %% Question: tu suggérais de reprendre ici "Ensuite, il y a la réduction de dimension données des refs (voir intro du papier JSTARS) et dans la réduction  de dimension, il  y a  la sélection de features." Le soucis que je vois, c'est que je détaille déjà dans l'intro les différentes méthodes de sélection/extraction et je trouve que ça devient pas très clair si je reprends tout en me basant sur le papier JSTARS. J'aurais dit avt "il y a trois types de sélection: filtre, wrapper, embedded blablabla" et ensuite ça me fait reprendre ici une autre distinction "supervisé/non-supervisé" si je me base sur jstars qui est un peu redondante il me semble. Qu'en dis-tu ?
    Second option is to reduce dimension. Features extraction/selection methods have been developed in order to reduce the dimension with various approaches described in Section~\ref{sec:intro}. In this study, this latter option is explored and a feature selection method named sequential forward features selection is presented after a brief overview of the selection algorithms in the next Section.

@@

        A measure of correct classification is based on an error matrix $M$, or \emph{confusion matrix}~\cite[Chapter 4]{congalton2008assessing}. An element $M_{ij}$ of the matrix represents the number of samples of class $i$ classified as class $j$. The confusion matrix allows the computation of several global and per-class indices related to the classification accuracy~\cite{congalton2008assessing}. Three global criteria were used :

@@

        The classification rate is  estimated by a $n_{cv}$- cross-validation ($n_{cv}$-CV)~\cite{opac-b1127878}.  To compute the $n_{cv}$-CV, a subset is removed  from $\mathcal{S}$ and  the GMM is learned with the remaining training samples. A test error is computed with the removed  training  samples  used  as  validation  samples.   The process is  iterated $n_{cv}$ times and  the estimated classification rate is computed as the mean  test error over the $n_{cv}$ subsets of $\mathcal{S}$.

@@

        \subsubsection{Similarity between distributions}
        The similarity  between two distributions is  quantified using
        divergence  measure~\cite{opac-b1097517}. Contrary  to measure
        of correct  classification, divergences are  computed directly
        on the trained model, with no need of validation samples.  Two
        particular   divergences   are   used  in   this   work:   the
        \emph{Kullback-Leibler}        divergence       and        the
        \emph{Jeffries-Matusita}  distance.  The  advantage  of  these
        divergences is  that they have  an explicit expression  in the
        case of Gaussian models. The  simplification allows to get rid
        of  any integration  calculus which  is a  major problem  when
        dealing with high-dimensional data.

@@

        The extension  to the multiclass  situation is done  by taking
        the weighted mean  of the KL divergences computed  on all pair
        of classes~\cite{bruzzone1995extension}:
        \begin{equation}
            C_{SKL} = \sum_{c=1}^{C} \sum_{c^\prime = c + 1}^{C} \pi_c \pi_{c^\prime} SKL_{cc^\prime}.
        \end{equation}

@@

        Table~\ref{tab:crit} summarizes the presented criterion functions and their characteristics.

@@

        The Sequential Forward Selection (SFS) starts with an empty set of selected features.  At each  step, the feature that most  improves the criterion function $J$ selected in Table~\ref{tab:crit} is added  to the set. This feature is definitively added to the pool of selected features and the algorithm stops when a given number of variables \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sfs} presents the process in details.

@@

        The SFFS works like the SFS but between each step of the SFS algorithm, a backward selection is operated to identify the less important feature. If the criterion value is higher than the best value ever obtained with a set of same size, the identified feature is picked out. The SBS step is repeated while removing the less important feature leads to an increase of the criterion value. Then SFS is called again. The algorithm stops when a given number of features \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sffs} provides details of the process.

        This SFFS algorithm eventually tests more solutions than the SFS algorithm. The results are expected to be better but the trade-off is an increased computational time which is dependent on the complexity of the dataset.

@@

\subsection{Statistical update rules}
        \subsubsection{Update for cross validation}
        \label{sec:update-cv}
        Based on \cite{fauvel2015fast}, a method to accelerate the $n_{cv}$-fold cross-validation process in the case of criterion functions based on correct classification measures was implemented. The idea is to estimate the GMM with the whole training set once and then, instead of training models on $(n_{cv}-1)$ folds, parameters of the complete model are used to derive those of sub-models.

>>>>

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Large-scale feature selection with Gaussian mixture models for remote sensing images}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

\IEEEPARstart{W}{ith} the increasing number of remote sensing missions, the quantity of available earth observation data for a given landscape becomes larger and larger. Satellite missions produce a huge amount of data on a regularly (daily) basis. From 2018, the EnMAP (Environmental Mapping and Analysis Program) satellites managed by the German space agency will produce images with 244 spectral bands, a spatial resolution of 30x30m per pixel and with a frequency of revisit of 4 days\cite{Müller09enmap}. The Hyperspectral Infrared Imager (HyspIRI) of NASA will also deliver images with 212 spectral bands. Additionally to hyperspectral data, the amount of available hypertemporal data increases a lot. For instance, the European satellites Sentinel-2 were launched successfully recently and 2 Terabytes of data will be released every days~\cite{drusch2012sentinel}. The Landsat open archive (\url{http://landsat.usgs.gov/products_data_at_no_charge.php}) has released thousands of images. This high volume of Earth observation data provides an accurate information of land state and functioning, and helps to improve the understanding of the planet~\cite{rs1010001}. However, processing such data is more and more challenging because of statistical and computational issues.

In the spectral or temporal domain, a pixel is represented by a vector
for   which  each   component  corresponds   to  a   spectral/temporal
measurement.   The size  of  the  vector is  therefore  the number  of
spectral  or temporal  measurements.  For  hyperspectral images,  this
number is  typically about several  hundreds while for  the Sentinel-2
multitemporal images, the number of spectro-temporal measurement for a
given  year  is approximately  one  thousand.   When working  in  high
dimensional  space,  statistical  methods  made for  low  or  moderate
dimensional  space do  not  adapt  well.  For  instance,  the rate  of
convergence of the statistical estimation decreases when the dimension
grows while conjointly the number of parameters to estimate increases,
making    the    estimation    of   the    model    parameters    very
difficult~\cite{donoho}.  Consequently,  with a limited  training set,
beyond a certain limit, the classification accuracy actually decreases
as the number of features increases~\cite{hughes}.  For the purpose of
classification,  these  problems are  related  to  the \emph{curse  of
  dimensionality}. This  is a  major drawback  in many  remote sensing
application since it  is difficult to collect a large  and an accurate
ground-truth.   An intensive  work has  been performed  in the  remote
sensing community  to build accurate classifiers  for high dimensional
images.  Bayesian models~\cite{book:landgrebe}, feature extraction and
feature   reduction   techniques~\cite{book:landgrebe,DR:guided:tour},
random   forest~\cite{1396322},  neural   networks~\cite{5411821}  and
kernel methods~\cite{kernel:methods:rs} have been investigated for the
classification of such images.

The volume of the data is  increasing dramatically with respect to the
number of measurement  per pixel. The data volume  of an hyperspectral
image  is  typically several  hundreds  of  Gigabytes per  acquisition
($\approx$  300km$^2$). Multitemporal  data are  now available  freely
from  internet  stream  (see  for instance  the  Copernicus  data  hub
\url{https://cophub.copernicus.eu/}).  This very  large volume of data
requires   specific   computing  infrastructure.    High   performance
computing   is   actually   investigated   by   the   remote   sensing
community~\cite{christophe2011remote,plaza2011high}.  Main  issues are
related to the use of parallel approach (multi-core, GPU, clusters) to
improve the  processing time, and  to the use of  streaming techniques
when data  does not fit in  memory.  One popular open  source software
solution  is  the Orfeo  Toolbox,  developed  by the  French  National
Agency~\cite{christophe2008orfeo}.   Streaming and  parallel computing
are  conveniently proposed  to the  users/developpers through  several
modules ``\emph{ready to use}''.

A method  to reduce  both statistical and  computational issues  is to
perform a reduction of the dimension. In fact, with the \emph{curse of
  dimensionality}     comes     the      \emph{blessing     of     the
  dimensionality}~\cite{bouveyron2014model}:  high   dimensional  data
space   exhibits  interesting   properties   for  the   classification
purpose.  In  particular,  it  is   possible  to  get  a  parsimonious
representation  of  the  data  while  maintaining  or  increasing  the
classification  accuracy~\cite{jimenez1998supervised}.  For  instance,
in land-cover  classification, given  a set  of spatial,  temporal and
spectral features, it is possible to  extract those which are the most
discriminant      for      the     purpose      of      classification
\cite{fassnacht2014comparison}.   In  hyperspectral   data,  from  the
hundreds of available spectral channels,  it is possible to reduce the
number of channels  to make the processing more efficient  in terms of
statistical complexity  and computational load. In  short, by reducing
the dimension better classification results is expected with a reduced
computational load.

@@

There  are  two  main  strategies  to  reduce  dimension~\cite[Chapter
1]{Guyon:2006:FEF:1208773}: features extraction and feature selection.
Feature extraction means reformulate  and summarize the information by
creating new features in combining  the existing ones, it is sometimes
referred to  \emph{feature construction}.   Linear combination  of the
initial features  can be extracted using  Principal Component Analysis
(PCA)~\cite{jimenez1998supervised}  or Independent  Component Analysis
\cite{villa2011hyperspectral}.  Supervised extraction  method has also
been investigated  such as Fischer discriminant  analysis and decision
boundary feature  extraction~\cite{book:landgrebe}.  To  the contrary,
features selection  extracts a subset of  existing features identified
as the  most relevant  by a  given criterion.   It has  the additional
advantage to be much more easily understandable for the end-user.

Features selection/extraction algorithms can be divided into three classes. The first class is called \emph{filter methods}. They select features independently of the classifier. Features are ranked according to some statistical measures, \emph{e.g.}, correlation or independence. For example,  PCA is a typical unsupervised filter method. Bruzzone \emph{et al.}\cite{bruzzone1995extension} develop a supervised filter method based on Jeffries-Matusita distance to maximize the separability of class distribution. Correlation between bands has been explored for feature selection in hyperspectral data~\cite{demir2008phase}. In general, these methods are fast and do not depend on any classifier. But they do not take into account the properties of the chosen classifier and do not optimize directly the classification accuracy.

The second class are known as \emph{wrapper methods}. They search for the best subset of variables for a given learning model. Since exhaustive searches are too expensive in terms of processing time, several sub-optimal search strategies have been designed, mainly iterative forward or backward search~\cite{whitney1971direct,marill1963effectiveness} or a combination of both~\cite{somol1999adaptive}. The advantage of such methods compare to filter methods is that they are dedicated to a particular model and to a particular learning problem. On the other hand, as they require the training of multiple models to test various set of variables, they are more time consuming.

@@

The third class corresponds to the \emph{embedded method}. They do not separate the features selection process from the learning algorithm and allow interaction between the two processes. A popular method embedded method is the \emph{Random Forest}. Embedded methods also exist for other models, \emph{e.g.} SVM \cite{guyon2002gene,weston2003use,tuia2015multiclass}.

Despite a large diversity of methods, feature selection algorithms usually do not scale well with the number of pixels to be processed~\cite{fauvel2015fast}. The training computational load is too important to compensate the reduced prediction computational load. Hence, feature selection is not widely used in operational situation. However, methods based on Gaussian Mixture Models (GMM) have several interesting properties that make them suitable for feature selection in the context of large amount of data. By taking advantage of their intrinsic properties, it is possible to increase the computational efficiency with respect to standard implementation.

The contribution of the paper is a forward feature selection method using GMM in continuation of~\cite{fauvel2015fast}. An efficient implementation of the method, using linear algebra on block matrix, is presented in order to handle large amount of data. Furthermore, a floating version of the algorithm is proposed. Several criteria are proposed to handle unbalanced training set.  Finally, the developed algorithm is made available as a remote module of the Orfeo Toolbox~\cite{christophe2008orfeo}.

@@

The remaining of the article is organized as follows. Section~\ref{sec:gmm-hd} presents GMM classifiers and the problems related to high-dimensional feature space. The features selection methods are detailed in Section~\ref{sec:selection}. Then an efficient implementation of the proposed selection method is presented in Section~\ref{sec:implementation}. Experimental results on two real high dimensional data set are given Section~\ref{sec:test}. Conclusion and perspectives conclude the paper in Section~\ref{sec:conclusion}.

@@

The    following    notations    are   used    in    the    remaining.
$\mathcal{S}  = \{\mathbf{x}_i,y_i\}_{i=1}^{n}$  denotes the  training
set where $\mathbf{x}_i \in \mathbb{R}^d$ is the vector of features of
the  $i^{th}$ sample,  $d$  the number  of spectral/temporal  features,
$y_i = 1,...,C$ the associated label, $C$ the total number of classes,
$n$ the  number of samples  and $n_c$ the  number of samples  of class
$c$.

@@

    For mixture models, it is assumed  that a given sample $\mathbf{x}$ is
    the realization  of a  random vector which  distribution is  a mixture
    (convex     combination)     of      several     class     conditioned
    distribution~\cite{Fraley00model-basedclustering}:
    \begin{align}
        p(\mathbf{x}) = \sum_{c=1}^{C} \pi_c f_c(\mathbf{x}|\theta)
    \end{align}
    where $\pi_c$ is  the prior, \emph{i.e.}, the  proportion of class
    $c$  and  $f_c$  a  parametric density  function  control by
    $\theta$.

    Among the possible parametric model,  the Gaussian one is the most
    used~\cite{bouveyron2014model}.   It assumes  that each  $f_c$ is,
    conditionally  to  $c$,  a  Gaussian  distribution  of  parameters
    $\boldsymbol{\mu}_c$    and    $\boldsymbol{\Sigma}_c$:
    \begin{multline}
        f_c(\mathbf{x}|\boldsymbol{\mu}_c, \boldsymbol{\Sigma}_c) = \\ \frac{1}{(2\pi)^{\frac{d}{2}} |\boldsymbol{\Sigma}_c|^{\frac{1}{2}}} \exp \left( -\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \right).
    \end{multline}

    It  is  referred  to  as  Gaussian  mixture  model  (GMM).   In  a
    supervised    learning    framework,    the    class    parameters
    $\boldsymbol{\mu}_c$,   $\boldsymbol{\Sigma}_c$   and  the   prior
    $\pi_c$ are  usually estimated  through the  conventional unbiased
    empirical estimators:
    \begin{align}
        \hat{\pi}_c &= \frac{n_c}{n},\\
        \hat{\boldsymbol{\mu}}_c &= \frac{1}{n_c} \sum_{\{i|y_i = c\}} \mathbf{x}_i ,\\
        \hat{\boldsymbol{\Sigma}}_c &= \frac{1}{(n_c - 1)} \sum_{\{i|y_i = c\}} (\mathbf{x}_i - \boldsymbol{\mu}_c) (\mathbf{x}_i - \boldsymbol{\mu}_c)^t.
    \end{align}
    To predict the  class of a new unseen sample,  the maximum \emph{a
      posteriori}  rule  is  used:
    \begin{equation*}
        \mathbf{x} \text{ belongs to } c \Leftrightarrow c = \text{arg} \max_{c \in C} p(c) p(\mathbf{x}|c).
    \end{equation*}
    Under the GMM,  and identifying $p(c)$ as $\pi_c$  and $p(\mathbf{x}|c)$ as
    $f_c(\mathbf{x}|\theta)$ and by taking the log, the decision function is obtained
    \begin{eqnarray}\label{eq:decision}
      Q_c(\mathbf{x}) &=& 2 \log \left( p(c) p(\mathbf{x}|c) \right) \nonumber \\
                      &=& - (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \nonumber \\
                      & &-\log (|\boldsymbol{\Sigma}_c|) + 2 \log (\pi_c) - d \log (2\pi).
    \end{eqnarray}

@@

    The computation of  eq.~(\ref{eq:decision}) requires the inversion
    of the covariance  matrix and the computation of  the logarithm of
    the determinant.  The  estimation of these terms  suffers from the
    curse of dimensionality\cite{bouveyron2014model}. In practice, the
    number of parameters $\eta_c$ to estimate for each class increases
    quadratically  with   respect  to  the  number   of  features,  as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation $n_c$  is small compared  to the number  of parameters
    $\eta_c$, the estimated covariance matrix is badly conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.  The  worst situation is  $n_c<\eta_c$ which leads  to a
    singular covariance matrix.  Unfortunately, this situation happens
    regularly in remote sensing.   For instance in hyperspectral image
    classification,  very few  labeled samples  are usually  available
    because of the difficulty and the cost to collect ground-truth.

@@

    %% Question: tu suggérais de reprendre ici "Ensuite, il y a la réduction de dimension données des refs (voir intro du papier JSTARS) et dans la réduction  de dimension, il  y a  la sélection de features." Le soucis que je vois, c'est que je détaille déjà dans l'intro les différentes méthodes de sélection/extraction et je trouve que ça devient pas très clair si je reprends tout en me basant sur le papier JSTARS. J'aurais dit avt "il y a trois types de sélection: filtre, wrapper, embedded blablabla" et ensuite ça me fait reprendre ici une autre distinction "supervisé/non-supervisé" si je me base sur jstars qui est un peu redondante il me semble. Qu'en dis-tu ?
    Second option is to reduce the dimension. Features extraction/selection methods have been developed in order to reduce the dimension with various approaches described in Section~\ref{sec:intro}. In this study, this latter option is explored and a feature selection method named sequential forward features selection is presented after an overview of the selection algorithms.

@@

        A measure of correct classification is based on an error matrix $M$, or \emph{confusion matrix}~\cite[Chapter 4]{congalton2008assessing}. The confusion matrix allows the computation of several global and per-class indices related to the classification accuracy~\cite{congalton2008assessing}. Three global criteria were used:

@@

        These  indices  are  estimated  from the  training  set  by  a
        $n_{cv}$- cross-validation ($n_{cv}$-CV)~\cite{opac-b1127878}.
        To  compute   the  $n_{cv}$-CV,  a  subset   is  removed  from
        $\mathcal{S}$  and  the  GMM  is learned  with  the  remaining
        training samples.  A test error  is computed with  the removed
        training samples  used as validation samples.   The process is
        iterated $n_{cv}$ times and  the estimated classification rate
        is computed as  the mean test error over  the $n_{cv}$ subsets
        of $\mathcal{S}$.

@@

        \subsubsection{Similarity between distributions}
        The similarity  between two distributions is  quantified using
        divergence  measure~\cite{opac-b1097517}. Contrary  to measure
        of correct  classification, divergences are  computed directly
        on  the  trained  model,  with  no  need  of  cross-validation
        estimation.  Two particular divergences are used in this work:
        the     \emph{Kullback-Leibler}     divergence     and     the
        \emph{Jeffries-Matusita}  distance.   The advantage  of  these
        divergences is  that they have  an explicit expression  in the
        case of Gaussian models. The  simplification allows to get rid
        of  any integration  calculus which  is a  major problem  when
        dealing with high-dimensional data.

@@

        The extension  to the multiclass  problem is done  by taking
        the weighted mean  of the KL divergences computed  on all pair
        of classes~\cite{bruzzone1995extension}:
        \begin{equation}
            C_{SKL} = \sum_{c=1}^{C} \sum_{c^\prime = c + 1}^{C} \pi_c \pi_{c^\prime} SKL_{cc^\prime}.
        \end{equation}

@@

        Table~\ref{tab:crit} summarizes the presented criterion functions and their characteristics. In the following $J$ denotes one criterion from Table~\ref{tab:crit}.

@@

        The Sequential Forward Selection (SFS) starts with an empty set of selected features.  At each  step, the feature that most  improves the criterion function $J$  is added  to the set. This feature is definitively added to the pool of selected features and the algorithm stops when a given number of variables \emph{maxVarNb} has been reached. The Algorithm~\ref{alg:sfs} presents the process in details.

@@

        The SFFS works as the SFS but between each step of the SFS algorithm, a backward selection is operated to identify the less important feature. If the criterion value is higher than the best value ever obtained with a set of same size, the identified feature is picked out. The SBS step is repeated while removing the less important feature leads to an increase of the criterion value. Then SFS is called again. The algorithm stops when a given number of features \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sffs} provides details of the process.

        This SFFS algorithm eventually evaluates more solutions than the SFS algorithm. The results are expected to be better but the trade-off is an increased computational time which is dependent on the complexity of the dataset.

@@

\subsection{Statistical update rules}
        \subsubsection{Update for cross validation}
        \label{sec:update-cv}
        Based on \cite{fauvel2015fast}, a method to accelerate the $n_{cv}$-fold cross-validation process in the case of criterion functions based on correct classification measures was implemented. The idea is to estimate the GMM with the whole training set once and then, instead of training models on $(n_{cv}-1)$ folds, parameters of the complete model are used to derive those of sub-models, thus reducing the whole complexity.
